---
title: "Machine Learning Assignment"
author: "Markus Gälli"
date: "2/18/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(caret)
library(e1071)
library(tidyr)
options(digits = 2)
```

## Overview

This document covers the analysis of excerise data collected from accelerometers from http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har

Different models were evaluated to maximize the prediction accuracy of the how the person exercised (correct, or one of 5 incorrect ways).

## Data
 
Data was downloaded from the assigned repository:
The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv



```{r LoadData}
setwd("~/R/Projects/Coursera Class/MachineLearningAssignment")
training <- read.csv("./data/pml-training.csv")
testing <- read.csv("./data/pml-testing.csv")
```

# Data Preparation
To facilitate the subsequent analysis, the training data has to be transformed somewhat:

- Turn the "classe" variable (indicating the exercise type) into a factor variable
- Put all the data into a data frame
- Make sure the data is imported as numeric (and not as character)
- Remove any variables with "NA" in them, as they can't be used for prediction
- Remove the first 7 columns as they do not contain data useful for this analysis

```{r DataPreproccessing}
training$classe <- factor(training$classe)
oldw <- getOption("warn")
options(warn = -1)
trainData <- as.data.frame(apply(training[,1:159],2,as.numeric))
options(warn = oldw)
trainData <- trainData %>% mutate(classe=training$classe)
cntNA <- apply(is.na(trainData), 2, sum)
ignore=which(cntNA != 0)
trainData <- trainData %>% select(-all_of(ignore)) %>% select(-(1:7))
```

The remaining data has the following dimension: `r nrow(trainData)` rows and `r ncol(trainData)` columns.

Next, we want to check for highly correlated variables. The diagonal has to be removed, and then a list is created with pairs of measurement variables that have a correlation of > 0.8. As the whole correlation matrix is included, each pair shows up twice.

```{r Correlations}
res <- cor(trainData[,-50])
high_corr <- which(abs(res)>0.8)
nc <- ncol(trainData)
dn <- seq(1,nc-1,by=1)
diagonal <- function(x){
    (x-1)*(length(x)+1)+1
}
diag <- diagonal(dn)
diag_ord <- high_corr %in% diag
remove <- high_corr[!diag_ord]
remove_row <- as.integer((remove-1)/49)+1
remove_col <- (remove-1) %% 49 +1

removeNames <- data.frame(rowN = names(trainData)[remove_row])
removeNames <- cbind(removeNames,colN = names(trainData)[remove_col])
removeNames
```

The highly correlated variables are removed from the data frame (not all of the above variables, but one from each pair).

```{r dataframeCleanup}
excludeNames <- c("gyros_forearm_z","accel_belt_y","accel_belt_z","gyros_dumbbell_x","gyros_arm_y","magnet_arm_x","magnet_arm_z",
                  "magnet_belt_x","pitch_dumbbell","yaw_dumbbell")

smallData <- trainData %>% select(-all_of(excludeNames))

```

This data set is now the one to work on.

## Modeling

First we break the data set into a training set and a testing set.

```{r trainAndTest}
set.seed(666)
inTrain <- createDataPartition(y=smallData$classe,p=0.7,list=FALSE)
trainingData <- smallData[inTrain,]
testingData <- smallData[-inTrain,]

```

The training set contains `r nrow(trainingData)` rows, and the testing set contains `r nrow(testingData)` rows. Both contain `r ncol(trainingData)` columns.

The following methods will be evaluated:

- Regression Trees
- Random Forests
- Boosting With Trees
- Support Vector Machines
- Linear Discriminant Analysis

For each method, a model is created using the training set, which is then used to predict on the training set. The results are evaluated using a confusion matrix.

```{r regressionTrees}
modFitTree <- train(classe~.,data=trainingData,method="rpart")
predTree <- predict(modFitTree,trainingData)
accTree<- confusionMatrix(predTree,trainingData$classe)

```

```{r regressionRF}
modFitRF <- train(classe ~ ., data=trainingData, method="rf",trControl = trainControl(method="cv"),number=3)
predRF <- predict(modFitRF,trainingData)
accRF <- confusionMatrix(predRF,trainingData$classe)
```

```{r regression Boosting}
modFitGBM <- train(classe ~ ., data=trainingData,method="gbm",verbose=FALSE)
predGBM <- predict(modFitGBM,trainingData)
accGBM <- confusionMatrix(predGBM, trainingData$classe)
```

```{r regression SVM}
modFitSVM <- svm(classe ~ ., data=trainingData)
predSVM <- predict(modFitSVM,trainingData)
accSVM <- confusionMatrix(predSVM,trainingData$classe)
```

```{R regressionLDA}
modFitLDA <- train(classe ~ ., data=trainingData, method= "lda")
predLDA <- predict(modFitLDA,trainingData)
accLDA <- confusionMatrix(predLDA,trainingData$classe)
```


The following graph shows the model accuracy for all four tested method.
```{r graph TrainingResults}
resModel <- c("Tree","RandomForest","GBM","LDA","SVM") 
resAcc <- c(accTree$overall[1],accRF$overall[1],accGBM$overall[1],accLDA$overall[1],accSVM$overall[1])
res <- data.frame(model=resModel,train_accuracy = resAcc)

g <- res %>% ggplot(aes(model,train_accuracy)) + geom_col() + 
    labs(title="Prediction Accuracies",x="Method",y="Accuracy") 
g
```

Random Forest, Support Vector Machines and Boosting WIth Trees clearly give the best answers, with high accuracy.
Now we validate the models by applying them to the testing data, which also provides the out of sample accuracy.

```{r CrossValidation}
resTree <- confusionMatrix(predict(modFitTree,testingData),testingData$classe)
resRF <- confusionMatrix(predict(modFitRF,testingData),testingData$classe)
resGBM <- confusionMatrix(predict(modFitGBM,testingData),testingData$classe)
resLDA <- confusionMatrix(predict(modFitLDA,testingData),testingData$classe)
resSVM <- confusionMatrix(predict(modFitSVM,testingData),testingData$classe)
resTest <- c(resTree$overall[1],resRF$overall[1],resGBM$overall[1],resLDA$overall[1],resSVM$overall[1])

res <- cbind(res,test_accuracy=resTest)
reslong <- res %>% gather("series","acc",-model)

g <- reslong %>% ggplot(aes(model,acc,fill=series)) + geom_col(position="dodge") + 
    labs(title="Prediction Accuracies",x="Model",y="Accuracy") + 
    scale_fill_discrete(name="Data Series",labels = c("Testing","Training"))
g
```

The graph shows both the in-sample and out-of-sample accuracies.



For the top 3 methods, the table below shows the accuracies:

Method        | In Model Accuracy | Out Of Model Accuracy
--------------|-------------------|----------------------
Random Forest | `r accRF$overall[1]` | `r resRF$overall[1]`
SVM           | `r accSVM$overall[1]`| `r resSVM$overall[1]`
Boosting with Trees | `r accGBM$overall[1]` | `r resGBM$overall[1]`

Based on these results, Random Forest seems to be the best method for the analysis of the provided data.
